Something about cetacean heart rates and ballistocardiograms

### 2021-08-27 analysis/paper/paper.Rmd
In progress: finding beats in manually identified motionless sections
Former-commit-id: ae905f37d61d7725792abc0f663843fd9b1eb82f

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
library(cetaceanbcg)
```

@@

# This window (in seconds) will be used for smoothing multiple signals and in
# heartbeat detection
window <- 0.25
prh_10hz <- readRDS(here::here("analysis/data/derived_data/prh_10hz.rds")) %>% 
  mutate(jerk_smooth = tma(prh_10hz$jerk, window * 10))
```

@@

```{r bcg-beats}
acc_400hz$surge_range <- local_range(acc_400hz$surge_filt, window * fs)
acc_400hz$bcg_beat <- find_beats(acc_400hz$surge_filt, window, fs)
acc_400hz$ecg_bpm[acc_400hz$ecg_beat] <- with(
  filter(acc_400hz, ecg_beat), 
  1 / as.numeric(lead(dt) - dt, unit = "mins")
)
acc_400hz$bcg_bpm[acc_400hz$bcg_beat] <- with(
  filter(acc_400hz, bcg_beat), 
  1 / as.numeric(lead(dt) - dt, unit = "mins")
)

@@

```{r bcg-results, fig.cap="A: The BCG signal is clearest in surge (longitudinal axis acceleration). Solid line: surge bandpass filtered to [1.0 25.0] Hz. Dashed line: local range of surge (LRS) in 0.25 s sliding window. Blue dots indicate peak of the BCG I-wave. B: Heart rates estimated from BCG (blue) match those from ECG (red). Temporal offset is due to both physiology and sensor clock differences. C: BCG signal is not as strong in heave (dorso-ventral axis) or sway (lateral axis) as in surge. The sway signal is likely due to movement of the chest cavity (see video S1). D: Some, but not all, heart beats are visible in the norm of the jerk vector."}
ca_tz <- "Etc/GMT+7"
bcg_begin <- lubridate::ymd_hms("2021-08-16 10:09:24.5", tz = ca_tz)
bcg_end <- lubridate::ymd_hms("2021-08-16 10:09:39", tz = ca_tz)
bcg_sample <- filter(acc_400hz, between(dt, bcg_begin, bcg_end))
bcg_sample_10hz <- filter(prh_10hz, between(dt, bcg_begin, bcg_end))

@@

# Fig 1a: surge and 
bcg_surge_lsr <- bcg_sample %>% 
  rename(Surge = surge_filt, LSR = surge_range) %>% 
  pivot_longer(c(Surge, LSR), 
               names_to = "var",
               values_to = "val")

@@

@@

@@

```{r demo-plot, fig.cap="A plot of random numbers"}
plot(rnorm(10))
```

Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot

```{r demo-inline-code}
x <- round(pi, 2)
```

>>>>

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
library(cetaceanbcg)
library(tidyverse)
```

@@

# This window (in seconds) will be used for smoothing multiple signals and in
# heartbeat detection
window <- 0.25
prh_10hz <- readRDS(here::here("analysis/data/derived_data/prh_10hz.rds")) %>% 
  mutate(jerk_smooth = tma(jerk, window * 10))
```

@@

```{r bcg-beats}
acc_400hz$surge_range <- local_range(acc_400hz$surge_filt, window * fs)
acc_400hz$bcg_beat <- find_beats_lsr(acc_400hz$surge_filt, window, fs)
acc_400hz$ecg_bpm[acc_400hz$ecg_beat] <- with(
  filter(acc_400hz, ecg_beat), 
  1 / as.numeric(lead(dt) - dt, unit = "mins")
)
acc_400hz$bcg_bpm[acc_400hz$bcg_beat] <- with(
  filter(acc_400hz, bcg_beat), 
  1 / as.numeric(lead(dt) - dt, unit = "mins")
)

@@

```{r corky-bcg, fig.cap="A: The BCG signal is clearest in surge (longitudinal axis acceleration). Solid line: surge bandpass filtered to [1.0 25.0] Hz. Dashed line: local range of surge in 0.25 s sliding window (smoothed). Blue dots indicate peak of the BCG I-wave. B: Heart rates estimated from BCG (blue) match those from ECG (red). Temporal offset is due to both physiology and sensor clock differences. C: BCG signal is not as strong in heave (dorso-ventral axis) or sway (lateral axis) as in surge. The sway signal is likely due to movement of the chest cavity (see video S1). D: Some, but not all, heart beats are visible as peaks in the norm of the jerk vector."}
ca_tz <- "Etc/GMT+7"
bcg_begin <- lubridate::ymd_hms("2021-08-16 10:09:24.5", tz = ca_tz)
bcg_end <- lubridate::ymd_hms("2021-08-16 10:09:39", tz = ca_tz)
bcg_sample <- filter(acc_400hz, between(dt, bcg_begin, bcg_end))
bcg_sample_10hz <- filter(prh_10hz, between(dt, bcg_begin, bcg_end))

@@

# A: Surge + LSR
bcg_surge_lsr <- bcg_sample %>% 
  rename(Surge = surge_filt, LSR = surge_range) %>% 
  pivot_longer(c(Surge, LSR), 
               names_to = "var",
               values_to = "val")

@@

# B: BPM

# C: Sway, heave

@@

# D: Jerk

@@

```{r bw-data}
fs <- 400
fny <- fs / 2
acc_filter <- signal::butter(4, c(1.5, 7.5) / fny, type = "pass") 
apply_filter <- function(x) signal::filtfilt(acc_filter, x)
window <- 1.8 #s
bw_400hz <- readRDS(here::here("analysis/data/derived_data/bw180905_53_400hz.rds")) %>% 
  mutate(across(c(surge, sway, heave), apply_filter, .names = "{.col}_filt"))

bw_10hz <- readRDS(here::here("analysis/data/derived_data/bw180905_53_10hz.rds")) %>% 
  mutate(jerk_smooth = tma(jerk, window * 10))

bw_elg <- readRDS(here::here("analysis/data/derived_data/bw180905_53_elg.rds"))
```

```{r bw-beats}
bw_400hz <- mutate(bw_400hz, surge_range = NA, bcg_beat = NA, bcg_bpm = NA)

for (row in seq(nrow(bw_eligible))) {
  i <- approx(bw_400hz$dt, 
              seq_along(bw_400hz$dt), 
              bw_elg[row, c("start", "stop")])$y
  i <- floor(i[1]):floor(i[2])
  region <- bw_400hz[i, ]
  surge <- region$surge_filt
  surge_range <- local_range(surge, window *fs)
  bcg_beat <- find_beats_lsr(surge, window, fs)
  bcg_bpm <- rep(NA, length(i))
  if (sum(bcg_beat) > 1) {
    bcg_dt <- region$dt[bcg_beat]
    bcg_bpm[bcg_beat] <- 1 / as.numeric(lead(bcg_dt) - bcg_dt, unit = "mins")
  }
  bw_400hz$surge_range[i] <- surge_range
  bw_400hz$bcg_beat[i] <- bcg_beat
  bw_400hz$bcg_bpm[i] <- bcg_bpm
}

bw_10hz$bcg_beat <- FALSE
bcg_idx <- approx(bw_10hz$dt, 
                  seq_along(bw_10hz$bcg_beat), 
                  bw_400hz$dt[bw_400hz$bcg_beat],
                  "constant")$y
bw_10hz$bcg_beat[bcg_idx] <- TRUE
```

```{r bw-bcg}
ca_tz <- "Etc/GMT+7"
bcg_begin <- lubridate::ymd_hms("2018-09-05 21:29:05", tz = ca_tz)
bcg_end <- lubridate::ymd_hms("2018-09-05 21:30:30", tz = ca_tz)
bcg_sample <- filter(bw_400hz, between(dt, bcg_begin, bcg_end))
bcg_sample_10hz <- filter(bw_10hz, between(dt, bcg_begin, bcg_end))

# A: Surge + LSR
bcg_surge_lsr <- bcg_sample %>% 
  rename(Surge = surge_filt, LSR = surge_range) %>% 
  pivot_longer(c(Surge, LSR), 
               names_to = "var",
               values_to = "val")

# B: BPM

p2 <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
  geom_line() +
  geom_point(color = "blue") +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "Heart rate (bpm)") +
  expand_limits(x = range(bcg_sample$dt), y = 0) +
  t

# C: Sway, heave

# D: Jerk

Figure \@ref(fig:corky-bcg) shows how we can have a caption and cross-reference for a plot.
